Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Word Processors

### 2018-10-02 main.tex
Implemented WERB Review Corrections by Marjorie McClain

\begin{document}
\maketitle
\begin{abstract}
\glsresetall
\noindent
\textbf{Purpose:} Modern mathematicians and scientists of math-related disciplines use \gls*{wp} to write and \gls*{cas} to calculate mathematical expressions. Usually, they translate the expressions manually between \gls*{wp} and \gls*{cas}. This process is time-consuming and error-prone. Our goal is to automate this translation. This paper uses \Maple{} and \Mathematica{} as the \gls*{cas}, and \LaTeX{} as our \gls*{wp}.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{wp} to write the paper and one or more \gls*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls*{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{wp} as well.} has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in{\mathbf R}$.
This mathematical expression is written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. One example of such a macro is given Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occur for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{}\footnote{The mention of
specific products, trademarks, or brand
names is for purposes of identification only. Such mention is not to be interpreted
in any way as an endorsement or certification of such products or brands by the
National Institute of Standards and Technology, nor does it imply that the products
so identified are necessarily the best available for the purpose. All trademarks
mentioned herein belong to their respective owners.} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\parencite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named to automatic \gls*{p2c} conversion~\parencite{POM-Tagger}.

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(0)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce a \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
as \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
as \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.
Note that partial results of this paper has been published in~\parencite{CICM:Paper}.

@@

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

@@

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. The goal of \sTeX{} was to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive amount of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is too complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not support.

@@

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care of each translation and provide details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. In consequence, translation~(\ref{eq:acot-alternatives-1}) solves the issue, as long as the user do not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z = -\iunit$.

Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
	\mathscr{W} \left\{ w_1(z), w_2(z) \right\} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation become unfeasible because the macro does not explicitely define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

The \Macro s do not garantee entirely disambiguated expressions. In~\Cref{tab:amb-latex} there are four examples of potentially ambiguous expressions. These expressions are unambiguous for the \LaTeX{} compiler since it only considers the very next token for superscripts and subscripts. Our translator follows the same rules to solve these issues.

@@

Another more questionable translation decision are alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently, depending on the order of the symbols. The reason is, that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' looks like indexing `$b$' by $4$. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons: (1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we define the following definitions

@@

In general, the translator is designed to find a work-around for disambiguating expressions. Only in case there is no way to solve the ambiguity with the defined rules, the translation process stops.

\section{The Translator}
All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as placeholders to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

These placeholders causes trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation*}
\verb|diff(f, [x$n])|,
\end{equation*}
where $f$ is an algebraic expression or an equation, $x$ is the name of the differentiation variable, and $n$ is an integer representing the $n$-th order differentiation\footnote{\url{https://www.maplesoft.com/support/help/maple/view.aspx?path=diff}, seen 07/2018}. A translation for $\frac{d^2x^2}{dx^2}$ should look like this
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation*}
but would end up as
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [xx])|.
\end{equation*}

@@

The \Macro s also allow one to specify optional arguments to distinguish between standard and another version of these functions. The Legendre and associated Legendre function of the first kind are examples of such cases. The library that defines translations for each macro uses the macro name as the primary key to identify the translations. The Legendre and associated Legendre function of the first kind both use the same macro \verb|\LegendreP|. To distinguish such cases, we use a special syntax for such cases, shown in~\Cref{tab:legendreP-lex}.

@@

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree, the \gls*{pom-pt}.

@@

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguate the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example is already interpreted as the double factorial function of $n$, since this notation is the standard interpretation in science. We wrote the second interpretation as the standard way in science to make it even more obvious. However, surprisingly, \Maple{} computes the first interpretation again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects like lists. This allows one to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three example for the \gls*{teo} list that groups some tokens.

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balancing of parentheses in expressions. An expression such as `$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use of interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

The \verb|SequenceTranslator| also handles positions of multiplication symbols. There are a couple of obvious choices to translate multiplication. The most common symbol for multiplications is still the white space (or no space between the tokens), as explained previously. Consider the simple expression `$2n\pi$'. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as `$f(x)$'. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.

>>>>

\begin{document}
\maketitle
\begin{abstract}
\glsresetall
\noindent
\textbf{Purpose:} Modern mathematicians and scientists of math-related disciplines use \glspl*{wp} to write and \glspl*{cas} to calculate mathematical expressions. Usually, they translate the expressions manually between \gls*{wp} and \gls*{cas}. This process is time-consuming and error-prone. Our goal is to automate this translation. This paper uses \Maple{} and \Mathematica{} as the \gls*{cas}, and \LaTeX{} as our \gls*{wp}.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{wp} to write the paper and one or more \glspl*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls*{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{wp} as well.} has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in{\mathbf R}$.
This mathematical expression is written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. One example of such a macro is given in Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occurs for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{}\footnote{The mention of
specific products, trademarks, or brand
names is for purposes of identification only. Such mention is not to be interpreted
in any way as an endorsement or certification of such products or brands by the
National Institute of Standards and Technology, nor does it imply that the products
so identified are necessarily the best available for the purpose. All trademarks
mentioned herein belong to their respective owners.} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (Figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\parencite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named as automatic \gls*{p2c} conversion~\parencite{POM-Tagger}.

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(1)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
\textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
is \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

@@

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition for what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

@@

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. The goal of \sTeX{} was to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive number of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is too complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not support.

@@

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care of each translation and provides details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. In consequence, translation~(\ref{eq:acot-alternatives-1}) solves the issue, as long as the user does not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z = -\iunit$.

Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
	\mathscr{W} \left\{ w_1(z), w_2(z) \right\} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitely define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

@@

The \Macro s do not guarantee entirely disambiguated expressions. In~\Cref{tab:amb-latex} there are four examples of potentially ambiguous expressions. These expressions are unambiguous for the \LaTeX{} compiler since it only considers the very next token for superscripts and subscripts. Our translator follows the same rules to solve these issues.

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently, depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' looks like indexing `$b$' by $4$. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons: (1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we make the following definitions

@@

In general, the translator is designed to find a work-around for disambiguating expressions. Only if there is no way to solve the ambiguity with the defined rules, the translation process stops.

\section{The Translator}
All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

These placeholders cause trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation*}
\verb|diff(f, [x$n])|,
\end{equation*}
where $f$ is an algebraic expression or an equation, $x$ is the name of the differentiation variable, and $n$ is an integer representing the $n$-th order differentiation\footnote{\url{https://www.maplesoft.com/support/help/maple/view.aspx?path=diff}, seen 07/2018}. A translation for $\frac{d^2x^2}{dx^2}$ should look like this
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation*}
but would end up as
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [xx])|.
\end{equation*}

@@

The \Macro s also allow one to specify optional arguments to distinguish between standard and another version of these functions. The Legendre and associated Legendre functions of the first kind are examples of such cases. The library that defines translations for each macro uses the macro name as the primary key to identify the translations. The Legendre and associated Legendre function of the first kind both use the same macro \verb|\LegendreP|. To distinguish such cases, we use a special syntax for such cases, shown in~\Cref{tab:legendreP-lex}.

@@

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example is already interpreted as the double factorial function of $n$, since this notation is the standard interpretation in science. We wrote the second interpretation as the standard way in science to make it even more obvious. However, surprisingly, \Maple{} computes the first interpretation again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

@@

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects like lists. This allows one to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three examples for the \gls*{teo} list that groups some tokens.

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balancing of parentheses in expressions. An expression such as `$(a]$' produces a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

The \verb|SequenceTranslator| also handles positions of multiplication symbols. There are a couple of obvious choices to translate multiplication. The most common symbol for multiplications is still the white space (or no space between the tokens), as explained previously. Consider the simple expression `$2n\pi$'. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as `$f(x)$'. Translating this to \Maple{} will give \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.
